---
title: "Data Cleaning and Preparation"
author: "James Thoburn"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = "")
```

## Load Libraries
```{r libraries}
library(tidyverse) # Allows us to easily manipulate and visualise data
```

## Load Data
```{r load-data}
listings <- read_csv("data/raw/listings.csv.gz")
```

## Column Selection

We start by examining the names of the different columns of the dataset to get a better understanding of what is and isn't relevant to our analysis.

```{r names}
names(listings)
```

`r ncol(listings)` columns is far too many for this analysis. As such, we reduce to only those that are plausibly relevant to predicting nightly price. A lot of the excluded columns either are related to how the data was scraped, rely on NLP to use or offer no predictive value towards the nightly price (e.g. `id` and `picture_url`). 

It should also be noted that `estimated_revenue_l365d` is excluded because it has a high risk of causing data leakage for the model since this column is directly related to the nightly price of the listing.

```{r column-selection}
listings <- listings |> 
  select(
    # Target variable
    price,
    
    # Property characteristics
    property_type,
    room_type,
    accommodates,
    bathrooms_text,
    bedrooms,
    beds,
    amenities,
    
    # Location
    neighbourhood_cleansed,
    neighbourhood_group_cleansed,
    latitude,
    longitude,
    
    # Host characteristics
    host_since,
    host_response_time,
    host_response_rate,
    host_acceptance_rate,
    host_is_superhost,
    host_listings_count,
    host_identity_verified,
    
    # Reviews
    number_of_reviews,
    review_scores_rating,
    review_scores_cleanliness,
    review_scores_location,
    review_scores_value,
    reviews_per_month,
    
    # Availability and booking
    minimum_nights,
    availability_365,
    instant_bookable
  )
```

## Price Cleaning

The `price` column is our target variable. Before anything else, we need to verify its data type and clean it if necessary.

```{r price-type}
class(listings$price)
```

Price is initially stored as a character string with a dollar sign and comma formatting (e.g. `$1,000.00`). We want this to be numeric for linear regression, so we strip these characters and convert  it to be numeric.

```{r price-clean}
listings <- listings |> 
  mutate(price = as.numeric(str_replace_all(price, "[$,]", "")))
```

With `price` now numeric, we check for zeroes and missing values. A price of zero is almost certainly a data entry error rather than a genuinely free listing.

```{r price-check}
listings |> 
  summarise(
    missing = sum(is.na(price)),
    zero = sum(price == 0, na.rm = TRUE),
    negative = sum(price < 0, na.rm = TRUE)
  )
```

The check reveals `r sum(is.na(listings$price))` missing prices, representing approximately 36% of the dataset. No zero or negative values are present. We will remove rows containing a missing price.

```{r remove-missing-prices}
listings <- listings |> 
  filter(!is.na(price))
```

After removing missing prices, `r nrow(listings)` listings remain.

## Predictor Cleaning

Now we will check for missing values in the different predictor variables.

```{r column-missingness-check}
listings |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))
```

### Missing Value Treatment

From this we can decide what to do with each column. Our options for handling missing values are:

1. **Drop the column entirely.**
2. **Replace missing values with the column median.**
3. **Add an "Unknown" category.**
4. **Convert to numeric then replace with column median.**
5. **Drop all of the rows with missing values for a given column.**

We treat each columns missingness individually based on the nature of the missing data and the proportion affected. 

`neighbourhood_group_cleansed` is entirely missing and as such is dropped entirely.

```{r drop-column}
listings <- listings |> 
  select(-neighbourhood_group_cleansed)
```

For Review scores, approximately 22% of the remaining listings have missing values. Since trying to manipulate these review scores would be making up data, rows containing missing review scores are removed. 

The remaining host-level missingness (25 rows each) and `host_is_superhost` are a small enough proportion of the data such that removing rows is preferable.

```{r drop-rows}
listings <- listings |> 
  filter(
    # Reviews
    !is.na(review_scores_location),
    !is.na(review_scores_value),
    !is.na(review_scores_cleanliness),
    !is.na(review_scores_rating),
    !is.na(reviews_per_month),
    # Host details
    !is.na(host_is_superhost),
    !is.na(host_since),
    !is.na(host_response_time),
    !is.na(host_response_rate),
    !is.na(host_acceptance_rate),
    !is.na(host_listings_count),
    !is.na(host_identity_verified)
  )
```

`beds` and `bedrooms` will have their missing values replaced with the column median. `bathrooms_text` will need to be cleaned to get the numeric values of the number of bathrooms, of which we then compute the column median to replace missing values.

```{r column-median}
listings <- listings |> 
  mutate(
    beds = if_else(is.na(beds), median(beds, na.rm = TRUE), beds),
    bedrooms = if_else(is.na(bedrooms), median(bedrooms, na.rm = TRUE), bedrooms),
    bathrooms = case_when(
      str_detect(bathrooms_text, regex("half-bath", ignore_case = TRUE)) ~ 0.5,
      TRUE ~ as.numeric(str_extract(bathrooms_text, "[0-9]+\\.?[0-9]*"))
    )
  ) |> 
  select(-bathrooms_text) |> 
  mutate(
    bathrooms = if_else(is.na(bathrooms), median(bathrooms, na.rm = TRUE), bathrooms)
  )
```

`host_response_time` is (to be converted to) a categorical variable and receives an "Unknown" category whilst `host_response_rate` and `host_acceptance_rate` are stored as character strings with percentage signs on the end, and as such will be cleaned and have missing values replaced with the column median.

```{r host-cleaning}
listings <- listings |> 
  mutate(
    host_response_rate = as.numeric(str_replace(host_response_rate, "%", "")),
    host_acceptance_rate = as.numeric(str_replace(host_acceptance_rate, "%", ""))
  ) |> 
  mutate(
    host_response_time = if_else(host_response_time == "N/A", "unknown", host_response_time),
    host_response_time = as.factor(host_response_time),
    host_response_rate = if_else(is.na(host_response_rate), median(host_response_rate, na.rm = TRUE), host_response_rate),
    host_acceptance_rate = if_else(is.na(host_acceptance_rate), median(host_acceptance_rate, na.rm = TRUE), host_acceptance_rate)
  )
```

We now verify that there are no missing values in our dataset.

```{r verify-missing}
listings |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") |>
  filter(n_missing > 0) |>
  arrange(desc(n_missing))
```

### Other Data Cleaning

`host_since` records the date a host joined Airbnb. The raw date is of no use to us as a predictor, but the count of the number of days since the host became a host is useful as a predictor. This will be stored as `host_tenure_days`.

```{r, host-tenure}
listings <- listings |> 
  mutate(
    host_since = as.Date(host_since),
    host_tenure_days = as.numeric(as.Date("2025-09-14") - host_since) # The data comes from 14/09/2025
  ) |> 
  select(-host_since) # We don't need host_since anymore
```

The `amenities` column stores an array of strings. Rather than attempting to encode all of the amenities -- of which there are likely to be hundreds across the dataset -- we will extract binary flags for those most likely to influence price. However, first we should examine the proportion of each of the important amenities in the set of listings:

```{r amenity-proportions}
listings |> 
  mutate(amenities = str_to_lower(amenities)) |> 
  summarise(
    wifi = mean(str_detect(amenities, "wifi")),
    kitchen = mean(str_detect(amenities, "kitchen")),
    air_conditioning = mean(str_detect(amenities, "air conditioning")),
    heating = mean(str_detect(amenities, "heating")),
    washer = mean(str_detect(amenities, "washer")),
    dryer = mean(str_detect(amenities, "dryer")),
    tv = mean(str_detect(amenities, "\\btv\\b")),
    pool = mean(str_detect(amenities, "pool")),
    gym = mean(str_detect(amenities, "gym")),
    elevator = mean(str_detect(amenities, "elevator")),
    parking = mean(str_detect(amenities, "parking")),
    hot_tub = mean(str_detect(amenities, "hot tub")),
    ev_charger = mean(str_detect(amenities, "ev charger"))
  ) |> 
  pivot_longer(everything(), names_to = "amenity", values_to = "proportion") |> 
  arrange(proportion)
```

Amenities that are present in fewer than 3% of listings (pool, hot tub, EV charger) are to be excluded as they are too rare to be seen as useful predictors. Amenities that are present in more than 89% of the listings (wifi, heating, kitchen) are also to be excluded as they have near-zero variance and struggle to discriminate between listings.

```{r amenities}
listings <- listings |> 
  mutate(amenities = str_to_lower(amenities)) |> 
  mutate(
    has_gym = str_detect(amenities, "gym"),
    has_air_conditioning = str_detect(amenities, "air conditioning"),
    has_elevator = str_detect(amenities, "elevator"),
    has_parking = str_detect(amenities, "parking"),
    has_tv = str_detect(amenities, "\\btv\\b"),
    has_dryer = str_detect(amenities, "dryer"),
    has_washer = str_detect(amenities, "washer")
  ) |> 
  select(-amenities)
```

Now we can simply save the cleaned dataset to `data/clean/` so that the subsequent sections of this analysis don't require us to re-run the cleaning pipeline to load.

```{r save}
write_csv(listings, "data/clean/listings_clean.csv")
```