---
title: "Exploratory Data Analysis"
author: "James Thoburn"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = "")
```

## Load Libraries
```{r libraries}
library(tidyverse) # Allows us to easily manipulate and visualise data
library(corrplot) # Allows us to visualise correlation matrices
library(patchwork) # Allows us to combine multiple plots into one layout easily
```

## Load Data
```{r load-data}
listings <- read_csv("data/clean/listings_clean.csv")
```

## Price Distribution

We begin by examining the distribution of our target variable, `price`, before any transformation is applied.

```{r price-summary}
listings |> 
  summarise(
    min = min(price),
    q1 = quantile(price, 0.25),
    median = median(price),
    mean = round(mean(price), 2),
    q3 = quantile(price, 0.75),
    max = max(price),
    sd = round(sd(price), 2)
  )
```

```{r price-distribution}
ggplot(listings, aes(x = price)) +
  geom_histogram(bins = 60, fill = "#4a90d9", colour = "white") +
  labs(
    title = "Distribution of Nightly Price",
    x = "Price ($)",
    y = "Count"
  ) +
  theme_minimal()
```

The distribution is severely right-skewed, with a maximum price of $74,100 pulling the histogram into an unreadable spike. The mean ($`r round(mean(listings$price), 2)`) sits well above the median ($`r median(listings$price)`), confirming the right skew of the nightly prices distribution. These extreme values are almost certainly either data-entry errors or entire-property long-term rental listings rather than genuine nightly prices.

We will examine the upper tail more closely before deciding on a capping threshold.

```{r examine-price-tail}
listings |>
  summarise(
    p90  = quantile(price, 0.90),
    p95  = quantile(price, 0.95),
    p99  = quantile(price, 0.99),
    p999 = quantile(price, 0.999)
  )
```

The upper tail reveals a very small number of extreme values -- the 99.9th percentile sits at approximately $4,621 yet the maximum is $74,100, suggesting these are not genuine nightly prices. As such, we will cap the price at the 99th percentile (approximately $908) to remove clear outliers while retaining legitimate high-end listings. This affects fewer than 1% of observations.

```{r price-cap}
cap <- quantile(listings$price, 0.99)

listings <- listings |> 
  filter(price <= cap)
```

With outliers removed, we apply a log transformation to price. Log transformation is appropriate here because price is strictly a positive quanity, the raw distribution of price is heavily right-skewed and proportional differences in price are more meaningful than absolute ones (e.g. a 50 dollar increase on a 100 dollar listing is far more significant than the same increase on a 500 dollar listing).

```{r log-price}
listings <- listings |> 
  mutate(log_price = log(price))
```

This log transformation gives us the following distribution:

```{r log-price-dist}
ggplot(listings, aes(x = log_price)) +
  geom_histogram(bins = 60, fill = "#4a90d9", colour = "white") +
  labs(
    title = "Distribution of Log Nightly Price",
    x = "Log Price",
    y = "Count"
  ) +
  theme_minimal()
```

Using `patchwork`, we can compare the two plots. 

```{r price-compare}
p1 <- ggplot(listings, aes(x = price)) +
  geom_histogram(bins = 60, fill = "#e05c5c", colour = "white") +
  labs(
    title = "Distribution of Nightly Price",
    x = "Price ($)",
    y = "Count"
  ) +
  theme_minimal()

p2 <- ggplot(listings, aes(x = log_price)) +
  geom_histogram(bins = 60, fill = "#4a90d9", colour = "white") +
  labs(
    title = "Distribution of Log Nightly Price",
    x = "Log Price",
    y = "Count"
  ) +
  theme_minimal()

p1 + p2 # Using patchwork library
```

The log transformation produces an approximately normal distribution, satisfying the normality assumption of the target variable and making the model more robust to the influence of high-value listings. As such, all subsequent analysis and modelling will use `log_price` as the target variable. The raw price is retained in the dataset so that we have good interpretability when reporting the final results.

## Predictor Distributions

Next, we examine the distribution of each predictor, grouped by type. Numeric predictors are visualised as histograms and categorical predictors as bar charts.

### Property Characteristics

First, we plot the numeric-typed property characteristics, namely `accomodates`, `bedrooms`, `beds` and `bathrooms`.

```{r property-numeric-dist}
p1 <- ggplot(listings, aes(x = accommodates)) +
  geom_histogram(bins = 20, fill = "#4a90d9", colour = "white") +
  labs(title = "Accommodates", x = NULL, y = "Count") +
  theme_minimal()

p2 <- ggplot(listings, aes(x = bedrooms)) +
  geom_histogram(bins = 20, fill = "#4a90d9", colour = "white") +
  labs(title = "Bedrooms", x = NULL, y = "Count") +
  theme_minimal()

p3 <- ggplot(listings, aes(x = beds)) +
  geom_histogram(bins = 20, fill = "#4a90d9", colour = "white") +
  labs(title = "Beds", x = NULL, y = "Count") +
  theme_minimal()

p4 <- ggplot(listings, aes(x = bathrooms)) +
  geom_histogram(bins = 20, fill = "#4a90d9", colour = "white") +
  labs(title = "Bathrooms", x = NULL, y = "Count") +
  theme_minimal()

(p1 + p2) / (p3 + p4)
```

`accomodates`, `bedrooms` and `beds` are all right-skewed, as expected. The majority of the listings accomodate a small group of people, have 1-2 bedrooms and 1-2 beds. There are some extreme values, but these are plausible for property listings designed for large groups and thus don't need to be removed. `bathrooms` is heavily concentrated at 1, notice the existence of 0.5 bathroom (half-bath only) listings.

Next, we plot `room_type`, the only categorical property characteristic.

```{r room-type-dist}
listings |> 
  count(room_type) |> 
  ggplot(aes(x = reorder(room_type, n), y = n)) +
  geom_col(fill = "#4a90d9") +
  coord_flip() +
  labs(title = "Room Type", x = NULL, y = "Count") +
  theme_minimal()
```

Room type is heavily dominated by entire home/apartment listings. Shared room and hotel room each represent a negligible portion of listings, and as such will produce unstable estimates if treated as separate categories. As such, we collapse these into an "Other" category.

```{r room-type-other-category}
listings <- listings |> 
  mutate(room_type = case_when(
    room_type %in% c("Shared room", "Hotel room") ~ "Other",
    TRUE ~ room_type
  ))
```

### Host Characteristics

```{r host-dist}
p1 <- ggplot(listings, aes(x = host_tenure_days)) +
  geom_histogram(bins = 40, fill = "#4a90d9", colour = "white") +
  labs(title = "Host Tenure (Days)", x = NULL, y = "Count") +
  theme_minimal()

p2 <- ggplot(listings, aes(x = host_listings_count)) +
  geom_histogram(bins = 40, fill = "#4a90d9", colour = "white") +
  labs(title = "Host Listings Count", x = NULL, y = "Count") +
  theme_minimal()

p3 <- ggplot(listings, aes(x = host_response_rate)) +
  geom_histogram(bins = 40, fill = "#4a90d9", colour = "white") +
  labs(title = "Host Response Rate", x = NULL, y = "Count") +
  theme_minimal()

p4 <- ggplot(listings, aes(x = host_acceptance_rate)) +
  geom_histogram(bins = 40, fill = "#4a90d9", colour = "white") +
  labs(title = "Host Acceptance Rate", x = NULL, y = "Count") +
  theme_minimal()

(p1 + p2) / (p3 + p4)
```

`host_tenure_days` is broadly distributed, reflecting Airbnb's growth since its foundation in 2008. The slight bimodal shape suggests that there were two waves of hosts joining the platform.

`host_listings_count` is severely right-skewed since the majority of hosts manage a single property, but a long tail of commercial operators manage hundreds to thousands of properties. This variable will require a log transformation prior to modelling.

```{r log-host-listings-count}
listings <- listings |> 
  mutate(log_host_listings_count = log1p(host_listings_count)) |> 
  select(-host_listings_count)
```

Both `host_response_rate` and `host_acceptance_rate` are heavily concentrated at 100%, reflecting that most hosts are active and maintain high engagement. The distributions are otherwise relatively flat.

### Reviews
```{r reviews-dist}
p1 <- ggplot(listings, aes(x = review_scores_rating)) +
  geom_histogram(bins = 30, fill = "#4a90d9", colour = "white") +
  labs(title = "Rating", x = NULL, y = "Count") +
  theme_minimal()

p2 <- ggplot(listings, aes(x = review_scores_cleanliness)) +
  geom_histogram(bins = 30, fill = "#4a90d9", colour = "white") +
  labs(title = "Cleanliness", x = NULL, y = "Count") +
  theme_minimal()

p3 <- ggplot(listings, aes(x = review_scores_location)) +
  geom_histogram(bins = 30, fill = "#4a90d9", colour = "white") +
  labs(title = "Location Score", x = NULL, y = "Count") +
  theme_minimal()

p4 <- ggplot(listings, aes(x = number_of_reviews)) +
  geom_histogram(bins = 40, fill = "#4a90d9", colour = "white") +
  labs(title = "Number of Reviews", x = NULL, y = "Count") +
  theme_minimal()

(p1 + p2) / (p3 + p4)
```

All three of Rating, Cleanliness and Location Score are heavily-left-skewed with a spike at 5.0. This is a well-known phenomenon in Airbnb called "review score inflation", where guests rarely leave scores less than a 4, making the scale effectively compressed into the 4.0-5.0 range. This has implications for modelling due to these variables having little variance in them. This signals that these three variables have a poor discriminatory power and is worth keeping in mind when interpreting their coefficients in the model.

Number of Reviews is severely right-skewed with the vast majority of the listings having fewer than 100 reviews and a long tail extending to near 2,000. This will need a log transformation prior to modelling, similar to `host_listings_count`.

```{r log-number-of-reviews}
listings <- listings |> 
  mutate(log_number_of_reviews = log1p(number_of_reviews)) |> 
  select(-number_of_reviews)
```

### Availability and Booking
```{r availability-dist}
p1 <- ggplot(listings, aes(x = availability_365)) +
  geom_histogram(bins = 40, fill = "#4a90d9", colour = "white") +
  labs(title = "Availability (365 Days)", x = NULL, y = "Count") +
  theme_minimal()

p2 <- ggplot(listings, aes(x = minimum_nights)) +
  geom_histogram(bins = 40, fill = "#4a90d9", colour = "white") +
  labs(title = "Minimum Nights", x = NULL, y = "Count") +
  theme_minimal()

p1 + p2
```

For Availability, we see spikes on both ends of the distribution. A large group of listings (0-100 days) are not commonly available, whilst another large group of listings (300-365 days) are almost always available. This likely reflects two different approaches that hosts may take to listing their property -- those who list rarely and those who list full-time. Whilst unusual, this distribution is valid and doesn't require any transformation.

The Minimum Nights distribution is heavily right-skewed with the vast majority of the listings requiring at least 1-7 nights and a tail extending absurdly to 1,000 nights. Values above 30 are almost certainly long-term rentals, as opposed to the usual short-term stays, which is outside the scope of this analysis. As such, we will introduce a cap and apply a log transformation.

```{r minimum-nights-cap}
listings <- listings |> 
  filter(minimum_nights <= 30)
```

```{r log-minimum-nights}
listings <- listings |> 
  mutate(log_minimum_nights = log1p(minimum_nights)) |> 
  select(-minimum_nights)
```

### Property Type

`property_type` contains `r listings |> count(property_type) |> nrow()` distinct categories, the majority of which have negligible counts. The top 9 categories cover 95% of the listings. Rather than retaining all 9, we collapse into four interpretable groups that align with the most meaningful distinctions in the London market.

```{r property-type-collapse}
listings <- listings |> 
  mutate(property_type = case_when(
    str_detect(property_type, regex("^Entire", ignore_case = TRUE)) ~ "Entire place",
    str_detect(property_type, regex("^Private room", ignore_case = TRUE)) ~ "Private room",
    str_detect(property_type, regex("^Room in hotel", ignore_case = TRUE)) ~ "Hotel room",
    TRUE ~ "Other"
  ))

# Check the property types
listings |> count(property_type) |> arrange(desc(n))
```

## Relationships with Price

We now examine how each predictor relates to `log_price`. Numeric predictors are visualised as scatterplots with a smooth trend line while categorical predictors are visualised as boxplots.

### Numeric Predictors

#### Property Characteristics

```{r scatter-property}
p1 <- ggplot(listings, aes(x = accommodates, y = log_price)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", colour = "#e05c5c") +
  labs(title = "Accommodates", x = NULL, y = "Log Price") +
  theme_minimal()

p2 <- ggplot(listings, aes(x = bedrooms, y = log_price)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", colour = "#e05c5c") +
  labs(title = "Bedrooms", x = NULL, y = "Log Price") +
  theme_minimal()

p3 <- ggplot(listings, aes(x = bathrooms, y = log_price)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", colour = "#e05c5c") +
  labs(title = "Bathrooms", x = NULL, y = "Log Price") +
  theme_minimal()

p4 <- ggplot(listings, aes(x = beds, y = log_price)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", colour = "#e05c5c") +
  labs(title = "Beds", x = NULL, y = "Log Price") +
  theme_minimal()

(p1 + p2) / (p3 + p4)
```

All four of the property size variables show a clear linear positive relationship with `log_price`. Accommodates, bedrooms, bathrooms and beds all increase consistently with price, which is to be expected since larger properties are known to command higher nightly rates. The relationships are reasonably linear on the log scale, suggesting that these predictors need not be transformed for our model. The vertical spread at each value reflects genuine price variation within properties of a given size, likely driven by location and quality differences which are captured by our other predictors.

#### Host Characteristics

```{r scatter-host}
p1 <- ggplot(listings, aes(x = host_tenure_days, y = log_price)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", colour = "#e05c5c") +
  labs(title = "Host Tenure (Days)", x = NULL, y = "Log Price") +
  theme_minimal()

p2 <- ggplot(listings, aes(x = log_host_listings_count, y = log_price)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", colour = "#e05c5c") +
  labs(title = "Log Host Listings Count", x = NULL, y = "Log Price") +
  theme_minimal()

p3 <- ggplot(listings, aes(x = host_response_rate, y = log_price)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", colour = "#e05c5c") +
  labs(title = "Host Response Rate", x = NULL, y = "Log Price") +
  theme_minimal()

p4 <- ggplot(listings, aes(x = host_acceptance_rate, y = log_price)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", colour = "#e05c5c") +
  labs(title = "Host Acceptance Rate", x = NULL, y = "Log Price") +
  theme_minimal()

(p1 + p2) / (p3 + p4)
```

Host tenure shows almost no relationship with log price as the trend line lies almost completely flat across the full range of tenure values. This suggests that how long someone has been hosting is not a meaningful price signal and that the variable may contribute very little to the model.

Log host listings show a weak positive relationship as hosts managing more properties tend to charge slightly higher prices, possibly suggesting that professional operators may manage higher-quality or better-located properties that command a higher price. The vertical striping reflects the discrete nature of the log-transformed count variable.

Both host response rate and host acceptance rate show near-flat trend lines with a wide vertical spread for every value. These variables appear to have very weak linear relationships with price. The severe clustering at 100% that we noticed earlier is again visible here as a dense vertical band on the right of both plots. These variables are unlikely to be strong predictors and their coefficients should be interpreted cautiously.

#### Reviews

```{r scatter-reviews}
p1 <- ggplot(listings, aes(x = review_scores_rating, y = log_price)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", colour = "#e05c5c") +
  labs(title = "Rating", x = NULL, y = "Log Price") +
  theme_minimal()

p2 <- ggplot(listings, aes(x = review_scores_cleanliness, y = log_price)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", colour = "#e05c5c") +
  labs(title = "Cleanliness", x = NULL, y = "Log Price") +
  theme_minimal()

p3 <- ggplot(listings, aes(x = review_scores_location, y = log_price)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", colour = "#e05c5c") +
  labs(title = "Location Score", x = NULL, y = "Log Price") +
  theme_minimal()

p4 <- ggplot(listings, aes(x = log_number_of_reviews, y = log_price)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", colour = "#e05c5c") +
  labs(title = "Log Number of Reviews", x = NULL, y = "Log Price") +
  theme_minimal()

(p1 + p2) / (p3 + p4)
```

Rating and cleanliness score both show almost flat lines with score inflation in the 4.0-5.0 section being made apparent yet again. These variables are unlikely to be meaningful predictors despite their intuitive appeal.

Location score shows a weak positive relationship where listings in higher-rated locations command slightly higher prices, which is to be expected for a city like London where neighbourhood prestige varies considerably. However, the relationship is weak and the same clustering issue as before arrives.

Log number of reviews shows a slight negative relationship with log price where higher priced listings tend to receive fewer reviews, likely due to them being less accessible (and thus, not booked as frequently) than a cheaper listing. The vertical striping again reflects the discrete nature of the underlying count variable before the log transformation.

Overall, review score variables are likely to be weak predictors due to score inflation. This is an important thing to understand when interpreting model results.

#### Availability

```{r scatter-availability}
p1 <- ggplot(listings, aes(x = availability_365, y = log_price)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", colour = "#e05c5c") +
  labs(title = "Availability (365 Days)", x = NULL, y = "Log Price") +
  theme_minimal()

p2 <- ggplot(listings, aes(x = log_minimum_nights, y = log_price)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "lm", colour = "#e05c5c") +
  labs(title = "Log Minimum Nights", x = NULL, y = "Log Price") +
  theme_minimal()

p1 + p2
```

Availability over 365 days shows a completely flat trend line with no discernible relationship with log price. The multimodal distribution noted earlier is visible here with vertical strips of visibly higher density. This variable is unlikely to contribute meaningfully to the model.

Log minimum nights shows a weak positive relationship where listings requiring longer minimum stays tend to charge slightly higher nightly prices, possibly reflecting that premium properties attract longer bookings. The vertical striping again reflects the discrete nature of minimum nights, even after the log transformation, as hosts tend to set values at integer numbers such as 1, 2, 3, 4, 5 and 7 nights.

### Categorical Predictors

#### Room and Property Type

```{r boxplot-categoricals}
p1 <- ggplot(listings, aes(x = room_type, y = log_price)) +
  geom_boxplot(fill = "#4a90d9", colour = "white") +
  coord_flip() +
  labs(title = "Room Type", x = NULL, y = "Log Price") +
  theme_minimal()

p2 <- ggplot(listings, aes(x = property_type, y = log_price)) +
  geom_boxplot(fill = "#4a90d9", colour = "white") +
  coord_flip() +
  labs(title = "Property Type", x = NULL, y = "Log Price") +
  theme_minimal()

p1 / p2
```

Room type shows a clear and meaningful separation in log price. Entire home/apartment listings command substantially more than private rooms. The "Other" category has a much wider interquartile range than the other two, reflecting the mixed composition of this category. This is likely to be one of the strongest predictors in the dataset and could contribute significantly to the model.

Property type mirrors this pattern closely as entire place listings price similarly to entire home/apartment in the room type plot, while private rooms are priced similarly here too. Hotel rooms sit closer to entire place pricing than private room pricing, which is to be expected. The "Other" category again has the widest interquartile range, again highlighting its mixed composition.

The strong similarity between room type and property type distributions suggests that these two variables are likely highly correlated with each other. This will be examined formally later when we carry out correlation analysis.

#### Host Details

```{r boxplot-host}
p1 <- ggplot(listings, aes(x = host_is_superhost, y = log_price)) +
  geom_boxplot(fill = "#4a90d9", alpha = 0.7) +
  labs(title = "Superhost", x = NULL, y = "Log Price") +
  theme_minimal()

p2 <- ggplot(listings, aes(x = host_identity_verified, y = log_price)) +
  geom_boxplot(fill = "#4a90d9", alpha = 0.7) +
  labs(title = "Identity Verified", x = NULL, y = "Log Price") +
  theme_minimal()

p3 <- ggplot(listings, aes(x = instant_bookable, y = log_price)) +
  geom_boxplot(fill = "#4a90d9", alpha = 0.7) +
  labs(title = "Instant Bookable", x = NULL, y = "Log Price") +
  theme_minimal()

p4 <- ggplot(listings, aes(x = host_response_time, y = log_price)) +
  geom_boxplot(fill = "#4a90d9", alpha = 0.7) +
  scale_x_discrete(labels = function(x) str_wrap(x, width = 10)) +
  labs(title = "Response Time", x = NULL, y = "Log Price") +
  theme_minimal()

(p1 + p2) / (p3 + p4)
```

Superhost status and identity verified show very little difference in log price distributions as the boxes are very similar in both position and spread. These variables are unlikely to be meaningful predictors of price, suggesting that guests do not systematically pay more for listings with these credentials.

Instant bookable shows a more noticeable difference, with non-instant bookable listings having a very slightly lower median price. The difference is negligible and this variable is unlikely to contribute strongly to the model.

Response time shows a slight trend. Listings where hosts respond within an hour tend to have higher median prices than those that take longer to reply. However, the interquartile ranges overlap substantially across all categories, suggesting that this is a weak signal at best. The "unkown" category sits in the middle, as would be expected.

Overall, host characteristic variables show surprisingly weak relationships with price across both numeric and categorical predictors. In turn, location and property size variables are likely to dominate the model.

## Correlation Analysis

We examine correlations between all numeric predictors to identify potential multicollinearity issues before modelling. High correlations between predictors can destabilise coefficient estimates and inflate standard errors.

```{r correlation-matrix}
numeric_predictors <- listings |> 
  select(where(is.numeric)) |> 
  select(-price, -log_price) |> 
  cor(use = "complete.obs")

corrplot(
  numeric_predictors,
  method = "color",
  type = "upper",
  tl.cex = 0.7,
  tl.col = "black",
  addCoef.col = "black",
  number.cex = 0.5,
  col = colorRampPalette(c("#e05c5c", "white", "#4a90d9"))(200),
  diag = FALSE
)
```

Severable notable correlations are present that warrant our attention before modelling.

**Property size variables:** `accomodates`, `bedrooms` and `beds` are highly intercorrelated (0.72-0.81). This is unsurprising since larger properties naturally have more bedrooms and beds, but including all three in the model simultaneously risks multicollinearity. This will be examined formally using Variance Inflation Factor (VIF) scores in the assumption section of this analysis.

**Review score variables:** `review_scores_rating`, `review_scores_cleanliness`, `review_scores_location` and `review_scores_value` are moderately to highly intercorrelated (0.61-0.90). This presents two problems -- weak signalling and high collinearity. Careful consideration will be given to which review score variables will be retained in the final model.

**Host response rate and review scores rating** show a moderate correlation (0.49), which tells us that more responsive hosts tend to receive higher review scores.

**Reviews per month and availability** show a moderate negative correlation (-0.23), which suggests that listings that are booked more frequently naturally have lower availability, which is a logical relationship rather than a multicollinearity concern.

All flagged correlations will be formally tested using VIF scores in the assumptions section. Variables with VIF above 5 will be considered for removal or consolidation.

## Saving Updated Dataset

The dataset has been further refined during EDA. We save the updated version to `data/clean/` for use in subsequent sections of this analysis.

```{r save}
write_csv(listings, "data/clean/listings_eda.csv")
```

---
*Previous: [Data Cleaning and Preparation](01_data.html) | Next: [Modelling](03_modelling.html)*