---
title: "Exploratory Data Analysis"
author: "James Thoburn"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = "")
```

## Load Libraries
```{r libraries}
library(tidyverse) # Allows us to easily manipulate and visualise data
library(corrplot) # Allows us to visualise correlation matrices
library(patchwork) # Allows us to combine multiple plots into one layout easily
```

## Load Data
```{r load-data}
listings <- read_csv("data/clean/listings_clean.csv")
```

## Price Distribution

We begin by examining the distribution of our target variable, `price`, before any transformation is applied.

```{r price-summary}
listings |> 
  summarise(
    min = min(price),
    q1 = quantile(price, 0.25),
    median = median(price),
    mean = round(mean(price), 2),
    q3 = quantile(price, 0.75),
    max = max(price),
    sd = round(sd(price), 2)
  )
```

```{r price-distribution}
ggplot(listings, aes(x = price)) +
  geom_histogram(bins = 60, fill = "#4a90d9", colour = "white") +
  labs(
    title = "Distribution of Nightly Price",
    x = "Price ($)",
    y = "Count"
  ) +
  theme_minimal()
```

The distribution is severely right-skewed, with a maximum price of $74,100 pulling the histogram into an unreadable spike. The mean ($`r round(mean(listings$price), 2)`) sits well above the median ($`r median(listings$price)`), confirming the right skew of the nightly prices distribution. These extreme values are almost certainly either data-entry errors or entire-property long-term rental listings rather than genuine nightly prices.

We will examine the upper tail more closely before deciding on a capping threshold.

```{r examine-price-tail}
listings |>
  summarise(
    p90  = quantile(price, 0.90),
    p95  = quantile(price, 0.95),
    p99  = quantile(price, 0.99),
    p999 = quantile(price, 0.999)
  )
```

The upper tail reveals a very small number of extreme values -- the 99.9th percentile sits at approximately $4,621 yet the maximum is $74,100, suggesting these are not genuine nightly prices. As such, we will cap the price at the 99th percentile (approximately $908) to remove clear outliers while retaining legitimate high-end listings. This affects fewer than 1% of observations.

```{r price-cap}
cap <- quantile(listings$price, 0.99)

listings <- listings |> 
  filter(price <= cap)
```

With outliers removed, we apply a log transformation to price. Log transformation is appropriate here because price is strictly a positive quanity, the raw distribution of price is heavily right-skewed and proportional differences in price are more meaningful than absolute ones (e.g. a 50 dollar increase on a 100 dollar listing is far more significant than the same increase on a 500 dollar listing).

```{r log-price}
listings <- listings |> 
  mutate(log_price = log(price))
```

This log transformation gives us the following distribution:

```{r log-price-dist}
ggplot(listings, aes(x = log_price)) +
  geom_histogram(bins = 60, fill = "#4a90d9", colour = "white") +
  labs(
    title = "Distribution of Log Nightly Price",
    x = "Log Price",
    y = "Count"
  ) +
  theme_minimal()
```

Using `patchwork`, we can compare the two plots. 

```{r price-compare}
p1 <- ggplot(listings, aes(x = price)) +
  geom_histogram(bins = 60, fill = "#e05c5c", colour = "white") +
  labs(
    title = "Distribution of Nightly Price",
    x = "Price ($)",
    y = "Count"
  ) +
  theme_minimal()

p2 <- ggplot(listings, aes(x = log_price)) +
  geom_histogram(bins = 60, fill = "#4a90d9", colour = "white") +
  labs(
    title = "Distribution of Log Nightly Price",
    x = "Log Price",
    y = "Count"
  ) +
  theme_minimal()

p1 + p2 # Using patchwork library
```

The log transformation produces an approximately normal distribution, satisfying the normality assumption of the target variable and making the model more robust to the influence of high-value listings. As such, all subsequent analysis and modelling will use `log_price` as the target variable. The raw price is retained in the dataset so that we have good interpretability when reporting the final results.

## Predictor Distributions

Next, we examine the distribution of each predictor, grouped by type. Numeric predictors are visualised as histograms and categorical predictors as bar charts.

### Property Characteristics

First, we plot the numeric-typed property characteristics, namely `accomodates`, `bedrooms`, `beds` and `bathrooms`.

```{r property-numeric-dist}
p1 <- ggplot(listings, aes(x = accommodates)) +
  geom_histogram(bins = 20, fill = "#4a90d9", colour = "white") +
  labs(title = "Accommodates", x = NULL, y = "Count") +
  theme_minimal()

p2 <- ggplot(listings, aes(x = bedrooms)) +
  geom_histogram(bins = 20, fill = "#4a90d9", colour = "white") +
  labs(title = "Bedrooms", x = NULL, y = "Count") +
  theme_minimal()

p3 <- ggplot(listings, aes(x = beds)) +
  geom_histogram(bins = 20, fill = "#4a90d9", colour = "white") +
  labs(title = "Beds", x = NULL, y = "Count") +
  theme_minimal()

p4 <- ggplot(listings, aes(x = bathrooms)) +
  geom_histogram(bins = 20, fill = "#4a90d9", colour = "white") +
  labs(title = "Bathrooms", x = NULL, y = "Count") +
  theme_minimal()

(p1 + p2) / (p3 + p4)
```

`accomodates`, `bedrooms` and `beds` are all right-skewed, as expected. The majority of the listings accomodate a small group of people, have 1-2 bedrooms and 1-2 beds. There are some extreme values, but these are plausible for property listings designed for large groups and thus don't need to be removed. `bathrooms` is heavily concentrated at 1, notice the existence of 0.5 bathroom (half-bath only) listings.

Next, we plot `room_type`, the only categorical property characteristic.

```{r room-type-dist}
listings |> 
  count(room_type) |> 
  ggplot(aes(x = reorder(room_type, n), y = n)) +
  geom_col(fill = "#4a90d9") +
  coord_flip() +
  labs(title = "Room Type", x = NULL, y = "Count") +
  theme_minimal()
```

Room type is heavily dominated by entire home/apartment listings. Shared room and hotel room each represent a negligible portion of listings, and as such will produce unstable estimates if treated as separate categories. As such, we collapse these into an "Other" category.

```{r room-type-other-category}
listings <- listings |> 
  mutate(room_type = case_when(
    room_type %in% c("Shared room", "Hotel room") ~ "Other",
    TRUE ~ room_type
  ))
```

### Host Characteristics

```{r host-dist}
p1 <- ggplot(listings, aes(x = host_tenure_days)) +
  geom_histogram(bins = 40, fill = "#4a90d9", colour = "white") +
  labs(title = "Host Tenure (Days)", x = NULL, y = "Count") +
  theme_minimal()

p2 <- ggplot(listings, aes(x = host_listings_count)) +
  geom_histogram(bins = 40, fill = "#4a90d9", colour = "white") +
  labs(title = "Host Listings Count", x = NULL, y = "Count") +
  theme_minimal()

p3 <- ggplot(listings, aes(x = host_response_rate)) +
  geom_histogram(bins = 40, fill = "#4a90d9", colour = "white") +
  labs(title = "Host Response Rate", x = NULL, y = "Count") +
  theme_minimal()

p4 <- ggplot(listings, aes(x = host_acceptance_rate)) +
  geom_histogram(bins = 40, fill = "#4a90d9", colour = "white") +
  labs(title = "Host Acceptance Rate", x = NULL, y = "Count") +
  theme_minimal()

(p1 + p2) / (p3 + p4)
```

`host_tenure_days` is broadly distributed, reflecting Airbnb's growth since its foundation in 2008. The slight bimodal shape suggests that there were two waves of hosts joining the platform.

`host_listings_count` is severely right-skewed since the majority of hosts manage a single property, but a long tail of commercial operators manage hundreds to thousands of properties. This variable will require a log transformation prior to modelling.

```{r log-host-listings-count}
listings <- listings |> 
  mutate(log_host_listings_count = log1p(host_listings_count)) |> 
  select(-host_listings_count)
```

Both `host_response_rate` and `host_acceptance_rate` are heavily concentrated at 100%, reflecting that most hosts are active and maintain high engagement. The distributions are otherwise relatively flat.

### Reviews
```{r reviews-dist}
p1 <- ggplot(listings, aes(x = review_scores_rating)) +
  geom_histogram(bins = 30, fill = "#4a90d9", colour = "white") +
  labs(title = "Rating", x = NULL, y = "Count") +
  theme_minimal()

p2 <- ggplot(listings, aes(x = review_scores_cleanliness)) +
  geom_histogram(bins = 30, fill = "#4a90d9", colour = "white") +
  labs(title = "Cleanliness", x = NULL, y = "Count") +
  theme_minimal()

p3 <- ggplot(listings, aes(x = review_scores_location)) +
  geom_histogram(bins = 30, fill = "#4a90d9", colour = "white") +
  labs(title = "Location Score", x = NULL, y = "Count") +
  theme_minimal()

p4 <- ggplot(listings, aes(x = number_of_reviews)) +
  geom_histogram(bins = 40, fill = "#4a90d9", colour = "white") +
  labs(title = "Number of Reviews", x = NULL, y = "Count") +
  theme_minimal()

(p1 + p2) / (p3 + p4)
```

All three of Rating, Cleanliness and Location Score are heavily-left-skewed with a spike at 5.0. This is a well-known phenomenon in Airbnb called "review score inflation", where guests rarely leave scores less than a 4, making the scale effectively compressed into the 4.0-5.0 range. This has implications for modelling due to these variables having little variance in them. This signals that these three variables have a poor discriminatory power and is worth keeping in mind when interpreting their coefficients in the model.

Number of Reviews is severely right-skewed with the vast majority of the listings having fewer than 100 reviews and a long tail extending to near 2,000. This will need a log transformation prior to modelling, similar to `host_listings_count`.

```{r log-number-of-reviews}
listings <- listings |> 
  mutate(log_number_of_reviews = log1p(number_of_reviews)) |> 
  select(-number_of_reviews)
```

### Availability and Booking
```{r availability-dist}
p1 <- ggplot(listings, aes(x = availability_365)) +
  geom_histogram(bins = 40, fill = "#4a90d9", colour = "white") +
  labs(title = "Availability (365 Days)", x = NULL, y = "Count") +
  theme_minimal()

p2 <- ggplot(listings, aes(x = minimum_nights)) +
  geom_histogram(bins = 40, fill = "#4a90d9", colour = "white") +
  labs(title = "Minimum Nights", x = NULL, y = "Count") +
  theme_minimal()

p1 + p2
```

For Availability, we see spikes on both ends of the distribution. A large group of listings (0-100 days) are not commonly available, whilst another large group of listings (300-365 days) are almost always available. This likely reflects two different approaches that hosts may take to listing their property -- those who list rarely and those who list full-time. Whilst unusual, this distribution is valid and doesn't require any transformation.

The Minimum Nights distribution is heavily right-skewed with the vast majority of the listings requiring at least 1-7 nights and a tail extending absurdly to 1,000 nights. Values above 30 are almost certainly long-term rentals, as opposed to the usual short-term stays, which is outside the scope of this analysis. As such, we will introduce a cap and apply a log transformation.

```{r minimum-nights-cap}
listings <- listings |> 
  filter(minimum_nights <= 30)
```

```{r log-minimum-nights}
listings <- listings |> 
  mutate(log_minimum_nights = log1p(minimum_nights)) |> 
  select(-minimum_nights)
```