---
title: "Exploratory Data Analysis"
author: "James Thoburn"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = "")
```

## Load Libraries
```{r libraries}
library(tidyverse) # Allows us to easily manipulate and visualise data
library(corrplot) # Allows us to visualise correlation matrices
```

## Load Data
```{r load-data}
listings <- read_csv("data/clean/listings_clean.csv")
```

## Price Distribution

We begin by examining the distribution of our target variable, `price`, before any transformation is applied.

```{r price-summary}
listings |> 
  summarise(
    min = min(price),
    q1 = quantile(price, 0.25),
    median = median(price),
    mean = round(mean(price), 2),
    q3 = quantile(price, 0.75),
    max = max(price),
    sd = round(sd(price), 2)
  )
```

```{r price-distribution}
ggplot(listings, aes(x = price)) +
  geom_histogram(bins = 100, fill = "#4a90d9", colour = "white") +
  labs(
    title = "Distribution of Nightly Price",
    x = "Price ($)",
    y = "Count"
  ) +
  theme_minimal()
```

The distribution is severely right-skewed, with a maximum price of $74,100 pulling the histogram into an unreadable spike. The mean ($`r round(mean(listings$price), 2)`) sits well above the median ($`r median(listings$price)`), confirming the right skew of the nightly prices distribution. These extreme values are almost certainly either data-entry errors or entire-property long-term rental listings rather than genuine nightly prices.

We will examine the upper tail more closely before deciding on a capping threshold.

```{r examine-price-tail}
listings |>
  summarise(
    p90  = quantile(price, 0.90),
    p95  = quantile(price, 0.95),
    p99  = quantile(price, 0.99),
    p999 = quantile(price, 0.999)
  )
```

The upper tail reveals a very small number of extreme values -- the 99.9th percentile sits at approximately $4,621 yet the maximum is $74,100, suggesting these are not genuine nightly prices. As such, we will cap the price at the 99th percentile (approximately $908) to remove clear outliers while retaining legitimate high-end listings. This affects fewer than 1% of observations.

```{r price-cap}
cap <- 908.00

listings <- listings |> 
  filter(price <= cap)
```

With outliers removed, we apply a log transformation to price. Log transformation is appropriate here because price is strictly a positive quanity, the raw distribution of price is heavily right-skewed and proportional differences in price are more meaningful than absolute ones (e.g. a 50 dollar increase on a 100 dollar listing is far more significant than the same increase on a 500 dollar listing).

```{r log-price}
listings <- listings |> 
  mutate(log_price = log(price))
```

This log transformation gives us the following distribution:

```{r log-price-dist}
ggplot(listings, aes(x = log_price)) +
  geom_histogram(bins = 100, fill = "#4a90d9", colour = "white") +
  labs(
    title = "Distribution of Log Nightly Price",
    x = "Log Price",
    y = "Count"
  ) +
  theme_minimal()
```
