---
title: "Modelling"
author: "James Thoburn"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment = "")
```

## Load Libraries
```{r libraries}
library(tidyverse)
library(caret) # Provides cross-validation and model training framework
library(glmnet) # Provides Ridge and Lasso regression
```

## Load Data
```{r load-data}
listings <- read_csv("data/clean/listings_eda.csv")
```

## Data Preparation

### Encoding Categorical Variables

Linear regression requires all predictors to be numeric. Categorical variables must be converted to dummy variables, which are essentially just binary (0 or 1) columns that represent each category. R's `lm()` function conveniently handles this automatically for factors, so we convert all character columns to factors first.

```{r convert-char-to-fact}
listings <- listings |> 
  mutate(across(where(is.character), as.factor))
```

We also convert logical columns to integers (0 or 1) rather than factors since they are already binary.

```{r convert-logi-to-fact}
listings <- listings |> 
  mutate(across(where(is.logical), as.integer))
```

### Removing Redundant Predictors

Based on the EDA findings we drop a small number of predictors before modelling. `latitude` and `longitude` are excluded since raw coordinates have no linear interpretable relationship with price, alongside the fact that neighbourhood is already captured by `neighbourhood_cleansed`. `review_scores_value` is excluded due to its high correlation with other review score variables noted in the correlation analysis.

```{r remove-predictors}
listings <- listings |> 
  select(-latitude, -longitude, -review_scores_value)
```

### Train/Test Split

We split the data into a training set (80%) and a test set (20%). The model is trained exclusively on the training set and evaluated on the test set. This gives us a solid estimate as to how the model performs on unseen data (i.e. new listings).

```{r train-test-split}
set.seed(0)

train_index <- createDataPartition(listings$log_price, p = 0.8, list = FALSE)

train <- listings[train_index,]
test <- listings[-train_index,]
```

`createDataPartition()` from the `caret` package splits the data whilst preserving the distribution of `log_price` across both sets, ensuring the split is representative. `set.seed(0)` ensures the split is reproducible and that anyone running this code will get an identical split.

## Baseline Model

From here we seek to produce a baseline model. The baseline model includes all remaining predictors with no refinement. The purpose of such a model is to establish a benchmark against which all following models can be compared. We train it on the training set only.

```{r baseline-model}
baseline <- lm(log_price ~ . - price, data = train)
summary(baseline)
```

We evaluate it on the test set using the Root Mean Square Error (RMSE) and R-squared. RMSE measures the average prediction error in log price units and R-squared measures the proportion of variance in log price explained by the model.

```{r baseline-evaluation}
baseline_predictions <- predict(baseline, newdata = test)

baseline_rmse <- sqrt(mean((test$log_price - baseline_predictions)^2))
baseline_r2 <- cor(test$log_price, baseline_predictions)^2

cat("Baseline RMSE:", round(baseline_rmse, 4), "\n")
cat("Baseline R-squared:", round(baseline_r2, 4), "\n")
```

## Refined Model

Based on the EDA findings, several predictors showed little to no relationship with log price and high intercorrelations with other predictors. These are removed before refitting to reduce noise and multicollinearity risk.

**Removed predictors and rationale:**

- `review_scores_rating` and `review_scores_cleanliness`: Highly correlated with other review scores (0.84 and 0.68) and low variance due to score inflation.
- `host_response_rate`: Weak price relationship and moderate correlation with review scores rating (0.49).
- `availability_365` and `host_tenure_days`: Near-flat trend lines against log price in EDA.
- `host_acceptance_rate`, `host_identity_verified`, `host_is_superhost`: Negligible price signal in categorical boxplots.

```{r refined-model}
refined <- lm(
  log_price ~ . - price
  - review_scores_rating
  - review_scores_cleanliness
  - host_response_rate
  - availability_365
  - host_tenure_days
  - host_acceptance_rate
  - host_identity_verified
  - host_is_superhost,
  data = train
)

summary(refined)
```

```{r refined-evaluation}
refined_predictions <- predict(refined, newdata = test)

refined_rmse <- sqrt(mean((test$log_price - refined_predictions)^2))
refined_r2 <- cor(test$log_price, refined_predictions)^2

cat("Refined RMSE:", round(refined_rmse, 4), "\n")
cat("Refined R-squared:", round(refined_r2, 4), "\n")
```

The refined model changes by approximately `r round(refined_r2/baseline_r2 - 1, 3) * 100`% of R-squared compared to the baseline, confirming to us that the removed predictors were very minor signals. The tradeoff is a simpler, more interpretable model with reduced multicollinearity risk. The small performance cost is acceptable since now we have a cleaner predictor set.

## Cross-Validated Model

Cross-validation gives a more reliable estimate of model performance than a single train/test split. In 10-fold cross validation, the training set is split into 10 equal folds. The model is trained 10 times, each time using 9 folds for training and the remaining fold for validation.

```{r cv-model}
cv_control <- trainControl(method = "cv", number = 10)

cv_model <- train(
  log_price ~ . - price
  - review_scores_rating
  - review_scores_cleanliness
  - host_response_rate
  - availability_365
  - host_tenure_days
  - host_acceptance_rate
  - host_identity_verified
  - host_is_superhost,
  data = train,
  method = "lm",
  trControl = cv_control
)

cv_model$results
```

```{r cv-evaluation}
cv_predictions <- predict(cv_model, newdata = test)

cv_rmse <- sqrt(mean((test$log_price - cv_predictions)^2))
cv_r2 <- cor(test$log_price, cv_predictions)^2

cat("CV Model RMSE:", round(cv_rmse, 4), "\n")
cat("CV Model R-squared:", round(cv_r2, 4), "\n")
```

The cross-validated model produces identical test set performance to the refined model (RMSE: `r round(cv_rmse,4)`, R-squared: `r round(cv_r2,4)`), which is expected since they use the same predictor set. Cross-validation changes how we estimate performance, not the model itself.

More importantly, the cross-validation results show a mean RMSE of `r round(cv_model$results["RMSE"], 4)` across 10 folds with a standard deviation of `r round(cv_model$results["RMSESD"], 4)`. This low variance across folds indicates the model is stable since performance is consistent regardless of which subset of the training data is used. The cross-validated R-squared of `r round(cv_model$results["Rsquared"], 4)` is very close to the test set R-squared of `r round(cv_r2, 4)`, indicating that the model generalises well to unseen data and is not overfitting to the training data.

## Lasso Model

Lasso regression extends ordinary least squares by adding a penalty term that shrinks predictor coefficients towards zero. Unlike standard regression, Lasso can shrink coefficients to exactly zero, performing automatic variable selection. The strength of the penalty is controlled by a parameter $\lambda$, selected here via cross-validation.

```{r lasso-prep}
# glmnet requires a model matrix rather than a formula
# model.matrix() converts factors to dummy variables automatically
# we remove the intercept column as glmnet adds its own

X_train <- model.matrix(log_price ~ . - price
                        - review_scores_rating
                        - review_scores_cleanliness
                        - host_response_rate
                        - availability_365
                        - host_tenure_days
                        - host_acceptance_rate
                        - host_identity_verified
                        - host_is_superhost,
                        data = train)[, -1]

X_test <- model.matrix(log_price ~ . - price
                        - review_scores_rating
                        - review_scores_cleanliness
                        - host_response_rate
                        - availability_365
                        - host_tenure_days
                        - host_acceptance_rate
                        - host_identity_verified
                        - host_is_superhost,
                        data = test)[, -1]

Y_train <- train$log_price
Y_test <- test$log_price
```

```{r lasso-cv}
set.seed(0)

lasso_cv <- cv.glmnet(X_train, Y_train, alpha = 1, nfolds = 10)

best_lambda <- lasso_cv$lambda.min
cat("Best lambda:", round(best_lambda, 6), "\n")

plot(lasso_cv)
```

```{r lasso-model}
lasso_model <- glmnet(X_train, Y_train, alpha = 1, lambda = best_lambda)
```

```{r lasso-evaluation}
lasso_predictions <- predict(lasso_model, s = best_lambda, newx = X_test)

lasso_rmse <- sqrt(mean((Y_test - cv_predictions)^2))
lasso_r2 <- cor(Y_test, lasso_predictions)^2

cat("Lasso Model RMSE:", round(lasso_rmse, 4), "\n")
cat("Lasso Model R-squared:", round(lasso_r2, 4), "\n")
```

```{r lasso-coefficients}
lasso_coefs <- coef(lasso_model, s = best_lambda)

lasso_coefs_df <- data.frame(
  predictor = rownames(lasso_coefs),
  coefficient = as.numeric(lasso_coefs)
) |> 
  filter(coefficient != 0, predictor != "(Intercept)") |> 
  arrange(desc(abs(coefficient)))

lasso_coefs_df
```

The Lasso model selects an optimal $\lambda$ of `r best_lambda` via cross-validation -- this is an extremely small penalty. As a result, `r nrow(lasso_coefs_df)` predictors retain non-zero coefficients and performance is virtually identical to the refined OLS model (RMSE: `r round(refined_rmse, 4)`, R-squared: `r round(refined_r2, 4)`). This indicates that after the manual variable removal carried out during EDA, little remaining redundancy exists for Lasso to eliminate. The result validates our earlier feature selection decisions and confirms that the predictors that we retained are all contributing meaningfully.

## Model Comparison

```{r model-comparison}
comparison <- data.frame(
  model = c("Baseline", "Refined", "Cross-Validated", "Lasso"),
  RMSE = c(baseline_rmse, refined_rmse, cv_rmse, lasso_rmse),
  R_squared = c(baseline_r2, refined_r2, cv_r2, lasso_r2)
) |> 
  mutate(RMSE = round(RMSE, 4),
         R_squared = round(R_squared, 4)
        )

comparison
```

```{r comparison-plot}
comparison |> 
  pivot_longer(cols = c(RMSE, R_squared),
               names_to = "metric",
               values_to = "value") |> 
  ggplot(aes(x = model, y = value, fill = model)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(title = "Model Comparison", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The cross-validated refined model is selected as the final model. It matches Lasso performance almost identically, is more interpretable than Lasso for communicating individual predictor effects and cross-validation confirms its stability across different subsets of the training data.

---
*Previous: [Exploratory Data Analysis](02_eda.html) | Next: [Assumptions](04_assumptions.html)*